Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2021-10-27 analysis/paper/paper.Rmd
preprint version

---
title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
author: |
 Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
date: |
 $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
output:
  bookdown::pdf_document2:
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
      #before_body: "../templates/latex/pre-print/before_body.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
    keep_tex: true
    #extra_dependencies: "subfig"
  bookdown::html_document2:
    df_print: kable
    keep_md: true
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
#linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
 Dental calculus has proven to contain a wealth of information 
 on the dietary habits of past populations. These insights have, to a large extent,
 been obtained by the extraction and identification of
 starch granules contained within the mineralised dental plaque from a wide range 
 of regions and time periods. The scope of previous studies have been limited to 
 microfossil extraction and identification to reconstruct dietary preferences from
 the archaeological record, and few studies have attempted to address the biases 
 of starch retention in dental calculus. Those that have considered this problem 
 have been limited to *in vivo* studies on modern humans and non-human primates.
 Here, we present a multispecies oral biofilm model, which allows experimental 
 research on  starch incorporation and retention to be conducted on *in vitro* 
 dental calculus in a controlled laboratory setting. The biofilms were exposed 
 to treatment solutions with known quantities of
 dietary starches (wheat and potato) during the 25-day growth period. After this,
 the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
 and counted. We show that the granule counts extracted from the model dental calculus 
 represented a low proportion (ranging from 0.06% to 0.16%) of the total number 
 of granules exposed to the biofilms throughout the experiment.
 Additionally, we found that the ratios of granule sizes from the extracted 
 starch granules differed from the original treatment solutions, with large 
 granules (>20 $\mu$m) consistently being under-represented.
 We also found a correlation
 between the absolute granule counts and dry-weight of the biofilm 
 (*r* = 0.66, 90%CI[0.46,0.79]), as well
 as between the concentration (count per mg) of granules and dry-weight 
 (*r* = 0.30, 90%CI[0.06,0.51]).  
 Our results reinforce previous *in vivo* studies suggesting that dental calculus 
 presents a very small, and partly biased picture of the original dietary intake 
 of starches, with an over-representation of plants producing granules smaller than
 20 $\mu$m in size. The experimental model presented here is well-suited to address
 the need for further validation of methods and biases associated with dietary
 research on dental calculus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/",
  fig.align = "center")
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
#cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
#cite_keys <- unlist(cite_keys, use.names = FALSE)
#rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T)
## end refs.bib ##

@@

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see \@ref(fig:protocol-fig) for an overview of the protocol).

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a 
Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose.

@@

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 $\tiny{M}$ ethylenediaminetetraacetic acid (EDTA) 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using 
the pipette to ensure that the counted transects were representative of the 
whole slide. The count from the slide was extrapolated to the whole sample
(see Supplementary Material).

@@

Statistical analysis was conducted in `r version$version.string`  [@R-base] and
the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
here [@R-here], and patchwork [@R-patchwork].
<!-- R packages used -->
A one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
as the grouping variable (GV) was conducted to explore the effect of the different 
starch treatments on biofilm growth.

To calculate size proportions, mean counts for each treatment were taken across 
both experimental plates for each treatment, resulting in a mean count for each
granule size category within each treatment.

@@

No $\alpha$-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
contained amylase activity that could be detected in the assay, ranging from
`r paste(signif(range(positive_amylase),3), collapse = " to ")` U/mL enzyme 
(Supplementary Material).
The results are not comparable to other studies presenting $\alpha$-amylase activity
levels in humans, as the unit definition may differ; 
however, they are sufficient to show that there is no activity in the system.

## Treatment type had no effect on biofilm growth

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot)A);
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. 
Further research is needed to determine why this differs from previous archaeological
findings.

@@

All scripts and data used in the analysis are available on OSF 
(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch),
following the format provided by the rrtools package [@R-rrtools].
More detailed protocols are available on OSF and protocols.io.

>>>>

---
title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
author: |
 Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
date: |
 $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
output:
  bookdown::html_document2:
    df_print: kable
    keep_md: true
  bookdown::pdf_document2:
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
    keep_tex: true
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
#linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
 Dental calculus has proven to contain a wealth of information 
 on the dietary habits of past populations. These insights have, to a large extent,
 been obtained by the extraction and identification of
 starch granules contained within the mineralised dental plaque from a wide range 
 of regions and time periods. The scope of previous studies have been limited to 
 microfossil extraction and identification to reconstruct dietary preferences from
 the archaeological record, and few studies have attempted to address the biases 
 of starch retention in dental calculus. Those that have considered this problem 
 have been limited to *in vivo* studies on modern humans and non-human primates.
 Here, we present a multispecies oral biofilm model, which allows experimental 
 research on  starch incorporation and retention to be conducted on *in vitro* 
 dental calculus in a controlled laboratory setting. The biofilms were exposed 
 to treatment solutions with known quantities of
 dietary starches (wheat and potato) during the 25-day growth period. After this,
 the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
 and counted. We show that the granule counts extracted from the model dental calculus 
 represented a low proportion (ranging from 0.06% to 0.16%) of the total number 
 of granules exposed to the biofilms throughout the experiment.
 Additionally, we found that the ratios of granule sizes from the extracted 
 starch granules differed from the original treatment solutions, with large 
 granules (>20 $\mu$m) consistently being under-represented.
 We also found a correlation
 between the absolute granule counts and dry-weight of the biofilm 
 (*r* = 0.66, 90%CI[0.46,0.79]), as well
 as between the concentration (count per mg) of granules and dry-weight 
 (*r* = 0.30, 90%CI[0.06,0.51]).  
 Our results reinforce previous *in vivo* studies suggesting that dental calculus 
 presents a very small, and partly biased picture of the original dietary intake 
 of starches, with an over-representation of plants producing granules smaller than
 20 $\mu$m in size. The experimental model presented here is well-suited to address
 the need for further validation of methods and biases associated with dietary
 research on dental calculus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.path = "../figures/",
  fig.align = "center")
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
cite_keys <- unlist(cite_keys, use.names = FALSE)
rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
## end refs.bib ##

@@

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationDentalPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).

@@

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing sterile artificial saliva (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a 
Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
The protocol is a slightly modified version of an Enzymatic Assay of $\alpha$-Amylase
(https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase) [@bernfeldAmylase1955], which measures the amount of 
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 mg of maltose in 6 minutes.

@@

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 $\tiny{M}$ ethylenediaminetetraacetic acid (EDTA) 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using 
the pipette to ensure that the counted transects were representative of the 
whole slide. The count from the slide was extrapolated to the whole sample
(see Supplementary Material for more detail).

@@

Statistical analysis was conducted in `r version$version.string`  [@R-base] and
the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
here [@R-here], and patchwork [@R-patchwork].
<!-- R packages used -->

To see if biofilm growth was differently affected by starch treatments, a 
one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
as the grouping variable (GV) was conducted.
To analyse granule counts and calculate size proportions, mean counts for each 
treatment were taken across
both experimental plates, resulting in a mean count for each
granule size category within each treatment.

@@

No $\alpha$-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
contained amylase activity that could be detected in the assay, ranging from
`r paste(signif(range(positive_amylase),3), collapse = " to ")` U/mL enzyme 
(full results can be found in the Supplementary Material).
The results are not comparable to other studies presenting $\alpha$-amylase activity
levels in humans, as the unit definition may differ; 
however, they are sufficient to show that there is no activity in the system.

## Treatment type had minimal effect on biofilm growth

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments, 
while medium granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot)A);
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeographyDrinking2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

All scripts and data used in the analysis are available on OSF 
(https://osf.io/uc5qy/) and Github (https://github.com/bbartholdy/byoc-starch),
following the format provided by the rrtools package [@R-rrtools].
More detailed protocols are available on OSF (https://osf.io/akevs/) and protocols.io (https://www.protocols.io/view/biofilm-growth-with-starch-treatment-bu7jnzkn).
Additional tables and figures are available in the Supplementary Material.
